\bf Looking at Non-Linear Dimension Reductions as Models in the Data Space

### 2024-05-20 paper.qmd
Added lwd plots

hex_grid_coloured_scurve1 <-  ggplot(
  data = hex_grid_with_counts_s_curve1, 
  aes(x = x, y = y)) + 
  geom_polygon(color = "black", aes(group = hex_poly_id, fill = std_counts)) +
  geom_point(data = umap_scurve_scaled, 
             aes(x = UMAP1, y = UMAP2), 
             alpha = 0.5,
             size = 0.5) + 
  scale_fill_viridis_c(direction = -1, na.value = "#ffffff", option = "C") #+
  #xlim(sc_xlims_b) + ylim(sc_ylims_b)

@@

hex_grid_coloured_scurve2 <-  ggplot(
  data = hex_grid_with_counts_s_curve2, 
  aes(x = x, y = y)) + 
  geom_polygon(color = "black", aes(group = hex_poly_id, fill = std_counts)) +
  geom_point(data = umap_scurve_scaled, 
             aes(x = UMAP1, y = UMAP2), 
             alpha = 0.5,
             size = 0.5) + 
  scale_fill_viridis_c(direction = -1, na.value = "#ffffff", option = "C") +
  xlim(sc_xlims) + ylim(sc_ylims)

@@

hex_grid_coloured_scurve3 <-  ggplot(
  data = hex_grid_with_counts_s_curve3, 
  aes(x = x, y = y)) + 
  geom_polygon(color = "black", aes(group = hex_poly_id, fill = std_counts)) +
  geom_point(data = umap_scurve_scaled, 
             aes(x = UMAP1, y = UMAP2), 
             alpha = 0.5,
             size = 0.5) + 
  scale_fill_viridis_c(direction = -1, na.value = "#ffffff", option = "C") +
  xlim(sc_xlims) + ylim(sc_ylims)

@@

Once setting up the hexagonal grid with an appropriate number of bins, some hexagonal bins may have few or no data points within them (see @fig-rmlowdenshex (a)). To ensure comprehensive coverage of the NLDR data, it is necessary to select hexagonal bins with a considerable number of data points. This involves calculating the number of points within each hexagon. Then, the standard count is computed by dividing the number of points within each hexagon by the maximum number of points in the grid (see @eq-equationp2). Next, bins with a standard count less than a certain benchmark value are removed (see @fig-rmlowdenshex (b)). The following steps will help in determining a suitable value for removing low-density hexagons:

1. Plot the distribution of the standardized counts (see @fig-stdctsScurve).
2. Examine the distribution of counts.
3. Select the first quantile value if the distribution is skewed.

<!--add distribution of density and add the first quartile as the benchmark-->

@@

df_bin_centroids_scurve <- scurve_model$df_bin_centroids
df_bin_scurve <- scurve_model$df_bin
```

```{r}
cell_count_scurve <- ggplot(df_bin_centroids_scurve, 
                            aes(x = reorder(as.factor(hexID), 
                                            -std_counts), 
                                y = std_counts)) + 
  geom_quasirandom() + 
  xlab("hexagonal id") + 
  ylab("standardized cell count") +
  geom_hline(yintercept = 0.2, color = "black", linewidth=0.8, alpha = 0.5) +
  theme(axis.title.x = element_text(size = 5),
        axis.title.y = element_text(size = 5, angle = 90),
        axis.text.x = element_text(size = 5, angle = 90),
        axis.text.y = element_text(size = 5))
```

@@

Furthermore, selecting the benchmark value for removing low-density hexagons is important. Removing unnecessary bins may lead to the formation of long edges and an uneven 2D model. Hence, rather than solely relying on the benchmark value to identify hexagons for removal, it's essential to consider the standard number of points in the neighboring hexagons of the identified low-density ones (see @fig-rmlowdenshex (c)). If neighboring bins also show low counts, only those bins will be removed. The remaining bins are used to construct the 2D model.

<!--add vis showing removing the hexagons using first quantile, and observing neighbors-->

The benchmark value for removing low-density hexagons ranges between $0$ and $1$. When analyzing how these benchmark values influence model performance, it's essential to observe the change in MSE as the benchmark value increases (see @fig-diagnosticpltScurvelwd). The MSE shows a gradual increase as the benchmark value progresses from $0$ to $1$. Evaluating this rate of increase is important. If the increment is not considerable, the decision might lean towards retaining low-density hexagons.

@@

<!--To generate errors for different total number of bins-->
```{r}
## To initialize benchmark values to remove low density hexagons
benchmark_rm_hex_vec <- append(seq(0, 0.99, by=0.1),0.99)

@@

benchmark_label_df <- tibble(x = 14, y = 0.6, 
                          label = "benchmark is 0.60.")

mse_scurve_lwd <- ggplot(error_rm_scurve, 
                     aes(x = b_non_empty, 
                         y = log(MSE))) + 
  geom_point(size = 1) +
  geom_line() + 
  geom_vline(xintercept = 15, linetype="solid", 
             color = "black", linewidth=0.8, alpha = 0.5) +
  geom_text(data = benchmark_label_df, 
            aes(x=x, 
                label= label, 
                y=y), 
            colour="black", 
            angle=90) +
  ylab("log(MSE)") +
  xlab("number of non-empty bins") +
  theme(axis.text.x = element_text(size = 5),
        axis.text.y = element_text(size = 5),
        axis.title.x = element_text(size = 5),
        axis.title.y = element_text(size = 5, angle = 90))
```

```{r}
#| fig-cap: "MSE from UMAP applied to S-curve dataset with diffferent bechmark choices. What is the effective bechmark value to remove low density hexagons? The residual plot have a steep slope at the beginning, indicating that a smaller bechmark value causes a larger amount of MSE. Then, the slope gradually declines or level off, indicating that a higher bechmark value generates a smaller MSE. Using the elbow method, it was observed that when the bechmark value is set to $0.6$, the lowest MSE occurred."
#| label: fig-mse-scurve-lwd
#| fig-pos: H
#| out-height: 50%

@@

>>>>

hex_grid_coloured_scurve1 <-  ggplot(
  data = hex_grid_with_counts_s_curve1, 
  aes(x = x, y = y)) + 
  geom_polygon(color = "black", aes(group = hex_poly_id, fill = std_counts)) +
  geom_point(data = umap_scurve_scaled, 
             aes(x = UMAP1, y = UMAP2), 
             alpha = 0.5,
             size = 0.5) + 
  scale_fill_viridis_c(direction = -1, na.value = "#ffffff", option = "C") +
  interior_annotation("a") #+
  #xlim(sc_xlims_b) + ylim(sc_ylims_b)

@@

hex_grid_coloured_scurve2 <-  ggplot(
  data = hex_grid_with_counts_s_curve2, 
  aes(x = x, y = y)) + 
  geom_polygon(color = "black", aes(group = hex_poly_id, fill = std_counts)) +
  geom_point(data = umap_scurve_scaled, 
             aes(x = UMAP1, y = UMAP2), 
             alpha = 0.5,
             size = 0.5) + 
  scale_fill_viridis_c(direction = -1, na.value = "#ffffff", option = "C") +
  xlim(sc_xlims) + ylim(sc_ylims) +
  interior_annotation("b")

@@

hex_grid_coloured_scurve3 <-  ggplot(
  data = hex_grid_with_counts_s_curve3, 
  aes(x = x, y = y)) + 
  geom_polygon(color = "black", aes(group = hex_poly_id, fill = std_counts)) +
  geom_point(data = umap_scurve_scaled, 
             aes(x = UMAP1, y = UMAP2), 
             alpha = 0.5,
             size = 0.5) + 
  scale_fill_viridis_c(direction = -1, na.value = "#ffffff", option = "C") +
  xlim(sc_xlims) + ylim(sc_ylims) +
  interior_annotation("c")

@@

Once setting up the hexagonal grid with an appropriate number of bins, some hexagonal bins may have few or no data points within them (see @fig-bins-scurve (b)). To ensure comprehensive coverage of the NLDR data, it is necessary to select hexagonal bins with a considerable number of data points. This involves calculating the number of points within each hexagon. Then, the standard count is computed by dividing the number of points within each hexagon by the maximum number of points in the grid. Next, bins with a standard count less than a benchmark value are removed (see @fig-lwd-scurve (a)). There is no specific rule for selecting a benchmark value. However, the following steps can help determine a suitable value for removing low-density hexagons:

1. Plot the distribution of the standardized counts (see @fig-stdcts-scurve).
2. Examine the distribution of counts.
3. Select the first quantile value if the distribution is skewed.

<!--add distribution of density and add the first quantile as the benchmark-->

@@

## To obtain the first quntile
benchmark1 <- round(quantile(df_bin_centroids_scurve$std_counts, names = FALSE)[2], 3)
```

```{r}
cell_count_scurve <- ggplot(df_bin_centroids_scurve, 
                            aes(x = reorder(as.factor(hexID), 
                                            -std_counts), 
                                y = std_counts)) + 
  geom_quasirandom() + 
  xlab("hexagonal id") + 
  ylab("standardized bin count") +
  geom_hline(yintercept = benchmark1, color = "black", linewidth=0.8, alpha = 0.5) +
  theme(axis.title.x = element_text(size = 5),
        axis.title.y = element_text(size = 5, angle = 90),
        axis.text.x = element_text(size = 5, angle = 90),
        axis.text.y = element_text(size = 5))
```

@@

The benchmark value for removing low-density hexagons ranges between $0$ and $1$. When analyzing how these benchmark values influence model performance, it's essential to observe the change in MSE as the benchmark value increases (see @fig-mse-scurve-lwd). The MSE shows a gradual increase as the benchmark value progresses from $0$ to $1$. Evaluating this rate of increase is important. If the increment is not considerable, the decision might lean towards retaining low-density hexagons.

@@

<!--To generate errors for different total number of bins-->
```{r}
## To initialize benchmark values to remove low density hexagons
benchmark_rm_hex_vec <- append(seq(0, 0.99, by=0.1), c(benchmark1, 0.99))

@@

benchmark_label_df <- tibble(x = 0.2, y = 0.1, 
                          label = paste0("benchmark is ", benchmark1))

mse_scurve_lwd <- ggplot(error_rm_scurve, 
                     aes(x = benchmark_rm_lwd, 
                         y = log(MSE))) + 
  geom_point(size = 1) +
  geom_line() + 
  geom_vline(xintercept = benchmark1, linetype="solid", 
             color = "black", linewidth=0.8, alpha = 0.5) +
  geom_text(data = benchmark_label_df, 
            aes(x=x, 
                label= label, 
                y=y), 
            colour="black", 
            angle=90) +
  ylab("log(MSE)") +
  xlab("benchmark value") +
  theme(axis.text.x = element_text(size = 5),
        axis.text.y = element_text(size = 5),
        axis.title.x = element_text(size = 5),
        axis.title.y = element_text(size = 5, angle = 90))
```

```{r}
#| fig-cap: "MSE from UMAP applied to S-curve dataset with diffferent bechmark choices. What is the effective bechmark value to remove low density hexagons? The residual plot have a steep slope at the end, indicating that a smaller bechmark value causes a small amount of MSE. Then, the slope gradually increases or level up, indicating that a higher bechmark value generates a higher MSE. Using the elbow method, it was observed that when the bechmark value is set to $0.242$, the lowest MSE occurred."
#| label: fig-mse-scurve-lwd
#| fig-pos: H
#| out-height: 50%

@@

Furthermore, selecting the benchmark value for removing low-density hexagons is important. Removing unnecessary bins may lead to the formation of long edges and an uneven \gD{} model. Hence, rather than solely relying on the benchmark value to identify hexagons for removal, it's essential to consider the standard number of points in the neighboring hexagons of the identified low-density bins (see @fig-lwd-scurve (b)). If neighboring bins also show low counts, only those bins will be removed. The remaining bins are used to construct the \gD{} model.

<!--add vis showing removing the hexagons using first quantile, and observing neighbors-->
<!--Using first quantile as the benchmark value-->
```{r}
hex_grid_with_low_counts <- full_join(hex_grid2, 
    df_bin_centroids_scurve, 
    by = c("hex_poly_id" = "hexID")) |>
  filter(!is.na(std_counts)) |>
  mutate(type = if_else(std_counts <= benchmark1, "low", "high")) |>
  filter(type == "low")

hex_grid_rm_lwd_scurve1 <-  ggplot(
  data = hex_grid_with_counts_s_curve2, 
  aes(x = x, y = y)) + 
  geom_polygon(color = "black", aes(group = hex_poly_id, fill = std_counts)) +
  geom_text(data = hex_grid_with_low_counts, 
            color = "black", 
            aes(x = c_x, y = c_y, group = hex_poly_id, label = "X", size = 1)) +
  scale_fill_viridis_c(direction = -1, na.value = "#ffffff", option = "C") +
  xlim(sc_xlims) + ylim(sc_ylims) +
  interior_annotation("a")
```

<!--Observing neighboring bins as well-->
```{r}
## First define low density bins using first quantile
df_bin_centroids_low <- df_bin_centroids_scurve |> 
  filter(std_counts <= benchmark1)

## Check neighboring bins
remove_id <- find_low_dens_hex(df_bin_centroids_all = df_bin_centroids_scurve, 
                               bin1 = effective_bin1_scurve, 
                               df_bin_centroids_low = df_bin_centroids_low)

## Remove the identified bins
df_bin_centroids_scurve_removed <- df_bin_centroids_scurve |>
  filter(hexID %in%remove_id)

hex_grid_with_low_counts_neighbor <- right_join(hex_grid, 
                                             df_bin_centroids_scurve_removed, 
                                             by = c("hex_poly_id" = "hexID"))

hex_grid_rm_lwd_scurve2 <-  ggplot(
  data = hex_grid_with_counts_s_curve2, 
  aes(x = x, y = y)) + 
  geom_polygon(color = "black", aes(group = hex_poly_id, fill = std_counts)) +
  geom_text(data = hex_grid_with_low_counts_neighbor, 
            color = "black", 
            aes(x = c_x, y = c_y, group = hex_poly_id, label = "X", size = 1)) +
  scale_fill_viridis_c(direction = -1, na.value = "#ffffff", option = "C") +
  xlim(sc_xlims) + ylim(sc_ylims) +
  interior_annotation("b")
```

```{r}
#| echo: false
#| label: fig-lwd-scurve
#| fig-pos: H
#| fig-cap: "(a) Identification of low-density hexagons using a benchmark value of $0.158$, and (c) Identification of low-density hexagons considering neighboring bins."
#| fig-width: 9
#| fig-height: 6

hex_grid_rm_lwd_scurve1 + hex_grid_rm_lwd_scurve2 +
  plot_layout(guides='collect', ncol = 2) &
  theme(legend.position='none', plot.tag = element_text(size = 8))
```

### 2024-05-20 paper.tex
Added lwd plots

Once setting up the hexagonal grid with an appropriate number of bins,
some hexagonal bins may have few or no data points within them (see
\textbf{?@fig-rmlowdenshex} (a)). To ensure comprehensive coverage of
the NLDR data, it is necessary to select hexagonal bins with a
considerable number of data points. This involves calculating the number
of points within each hexagon. Then, the standard count is computed by
dividing the number of points within each hexagon by the maximum number
of points in the grid (see \textbf{?@eq-equationp2}). Next, bins with a
standard count less than a certain benchmark value are removed (see
\textbf{?@fig-rmlowdenshex} (b)). The following steps will help in
determining a suitable value for removing low-density hexagons:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  Plot the distribution of the standardized counts (see
  \textbf{?@fig-stdctsScurve}).
\item
  Examine the distribution of counts.
\item
  Select the first quantile value if the distribution is skewed.
\end{enumerate}

@@

Furthermore, selecting the benchmark value for removing low-density
hexagons is important. Removing unnecessary bins may lead to the
formation of long edges and an uneven 2D model. Hence, rather than
solely relying on the benchmark value to identify hexagons for removal,
it's essential to consider the standard number of points in the
neighboring hexagons of the identified low-density ones (see
\textbf{?@fig-rmlowdenshex} (c)). If neighboring bins also show low
counts, only those bins will be removed. The remaining bins are used to
construct the 2D model.

The benchmark value for removing low-density hexagons ranges between
\(0\) and \(1\). When analyzing how these benchmark values influence
model performance, it's essential to observe the change in MSE as the
benchmark value increases (see \textbf{?@fig-diagnosticpltScurvelwd}).
The MSE shows a gradual increase as the benchmark value progresses from
\(0\) to \(1\). Evaluating this rate of increase is important. If the
increment is not considerable, the decision might lean towards retaining
low-density hexagons.

@@

\caption{\label{fig-mse-scurve-lwd}MSE from UMAP applied to S-curve
dataset with diffferent bechmark choices. What is the effective bechmark
value to remove low density hexagons? The residual plot have a steep
slope at the beginning, indicating that a smaller bechmark value causes
a larger amount of MSE. Then, the slope gradually declines or level off,
indicating that a higher bechmark value generates a smaller MSE. Using
the elbow method, it was observed that when the bechmark value is set to
\(0.6\), the lowest MSE occurred.}

>>>>

Once setting up the hexagonal grid with an appropriate number of bins,
some hexagonal bins may have few or no data points within them (see
Figure~\ref{fig-bins-scurve} (b)). To ensure comprehensive coverage of
the NLDR data, it is necessary to select hexagonal bins with a
considerable number of data points. This involves calculating the number
of points within each hexagon. Then, the standard count is computed by
dividing the number of points within each hexagon by the maximum number
of points in the grid. Next, bins with a standard count less than a
benchmark value are removed (see Figure~\ref{fig-lwd-scurve} (a)). There
is no specific rule for selecting a benchmark value. However, the
following steps can help determine a suitable value for removing
low-density hexagons:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  Plot the distribution of the standardized counts (see
  Figure~\ref{fig-stdcts-scurve}).
\item
  Examine the distribution of counts.
\item
  Select the first quantile value if the distribution is skewed.
\end{enumerate}

@@

The benchmark value for removing low-density hexagons ranges between
\(0\) and \(1\). When analyzing how these benchmark values influence
model performance, it's essential to observe the change in MSE as the
benchmark value increases (see Figure~\ref{fig-mse-scurve-lwd}). The MSE
shows a gradual increase as the benchmark value progresses from \(0\) to
\(1\). Evaluating this rate of increase is important. If the increment
is not considerable, the decision might lean towards retaining
low-density hexagons.

@@

\caption{\label{fig-mse-scurve-lwd}MSE from UMAP applied to S-curve
dataset with diffferent bechmark choices. What is the effective bechmark
value to remove low density hexagons? The residual plot have a steep
slope at the end, indicating that a smaller bechmark value causes a
small amount of MSE. Then, the slope gradually increases or level up,
indicating that a higher bechmark value generates a higher MSE. Using
the elbow method, it was observed that when the bechmark value is set to
\(0.242\), the lowest MSE occurred.}

Furthermore, selecting the benchmark value for removing low-density
hexagons is important. Removing unnecessary bins may lead to the
formation of long edges and an uneven \gD{} model. Hence, rather than
solely relying on the benchmark value to identify hexagons for removal,
it's essential to consider the standard number of points in the
neighboring hexagons of the identified low-density bins (see
Figure~\ref{fig-lwd-scurve} (b)). If neighboring bins also show low
counts, only those bins will be removed. The remaining bins are used to
construct the \gD{} model.

\includegraphics[width=1\textwidth,height=\textheight]{paper_files/figure-pdf/fig-lwd-scurve-1.pdf}

\caption{\label{fig-lwd-scurve}(a) Identification of low-density
hexagons using a benchmark value of \(0.158\), and (c) Identification of
low-density hexagons considering neighboring bins.}
